### 2023-05-19 proofs/tex/alg-subtyping-lemmas.tex
Add merge completeness

\begin{lemma}[Completeness of the Positive Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and
    there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
    Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us']]$
    and $[[Θ ⊢ us ⇒ us']]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
    Let us consider the last rule used in the derivation of  $[[ Γ ⊢ [us]uP ≥ iQ ]]$,
    but first consider the base case for the substitution $[[ [us]uP ]]$:

>>>>

\begin{lemma}[Completeness of the Positive Subtyping] \label{lemma:pos-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and
    there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
    Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us']]$
    and $[[Θ ⊢ us ⇒ us']]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
    Let us consider the last rule used in the derivation of  $[[ Γ ⊢ [us]uP ≥ iQ ]]$,
    but first consider the base case for the substitution $[[ [us]uP ]]$:

### 2023-05-19 proofs/tex/solution-merge-lemmas.tex
Add merge completeness

@@

\begin{lemma} [Completeness and Initiality of Solution Merge] 
    \label{lemma:merge-completeness}
    Suppose that $[[us1 : Θ1]]$ and $[[us2 : Θ2]]$
    and there exists $[[us]]$
    such that $[[Θ1 ⊢ us1 ⇒ us]]$ and $[[ Θ2 ⊢ us2 ⇒ us]]$.
    Then $[[us1 & us2]]$ is defined and 
    $[[ Θ1 ∪ Θ2 ⊢ us1 & us2 ⇒ us]]$.
\end{lemma}

\begin{lemma} [Solution Weakening is Monotonous]
    \label{lemma:weakening-monotonicity}
    if $[[us1 : Θ]]$ and $[[us2 : Θ]]$ and 
    $[[us1 ⊆ us2]]$ then $[[Θ ⊢ us2 ⇒ us1]]$.
\end{lemma}

>>>>

\begin{lemma} [Solution Weakening is Monotonous]
    \label{lemma:weakening-monotonicity}
    if $[[us1 : Θ1]]$, $[[us2 : Θ2]]$,
    and $[[us1 ⊆ us2]]$ then $[[Θ1 ⊢ us2 ⇒ us1]]$.
\end{lemma}
\begin{proof}
    We need to show that for any $[[usEntry]] \in [[us1]]$,
    there exists $[[usEntry']] \in [[us2]]$ such that 
    $[[Θ1 ⊢ usEntry' ⇒ usEntry]]$.
    Let us consider an arbitrary $[[usEntry]] \in [[us1]] \subseteq [[us2]]$
    restricting $[[α̂±]]$. 
    Notice that $[[Θ1(α̂±) ⊢ usEntry ]]$ since 
    $[[usEntry]] \in [[us1 : Θ1]]$.
    By \cref{lemma:entry-weakening-preorder}, $[[Θ1 ⊢ usEntry ⇒ usEntry]]$. 
\end{proof}

@@

\begin{lemma} [Completeness of Unification Solution Entry Merge]
    \label{lemma:unif-entry-merge-completeness}
    For a fixed context $[[Γ]]$,
    suppose that  $[[Γ ⊢ usEntry1]]$ and $[[Γ ⊢ usEntry2]]$, 
    where $[[Γ ⊢ usEntryi]]$ are equivalence-shaped restrictions.
    Suppose there exists an equivalence-shaped restriction 
    $[[Γ ⊢ usEntry]]$ such that $[[Γ ⊢ usEntry ⇒ usEntry1]]$ and $[[Γ ⊢ usEntry ⇒ usEntry2]]$.
    Then $[[usEntry1 & usEntry2]]$ is defined and
    $[[Γ ⊢ usEntry ⇒ usEntry1 & usEntry2]]$.
\end{lemma}
\begin{proof}
    By the soundness of unification solution entry merge (\cref{lemma:unif-sol-ent-merge-soundness}),
    Since the restrictions are equivalence-shaped, 
    the only possible rules inferring the weakening judgments are
    the positive \ruleref{\ottdruleSImpEPEqEqLabel} and the negative \ruleref{\ottdruleSImpENEqEqLabel}.
    Since $[[usEntry1]]$, $[[usEntry2]]$, and $[[usEntry]]$ 
    have the same polarity, we consider the positive case,
    and the negative case is proved analogously.

    Let us assume that $[[usEntry1]]$ is $[[pua :≈ iP1]]$,
    $[[usEntry2]]$ is $[[pua :≈ iP2]]$, $[[usEntry]]$ is $[[pua :≈ iQ]]$.
    Then by inversion of \ruleref{\ottdruleSImpEPEqEqLabel},
    \begin{itemize}
        \item $[[Γ ⊢ iQ ≈ iP1]]$, and hence $[[nf(iQ) = nf(iP1)]]$ by \cref{lemma:subt-equiv-algorithmization},
        \item $[[Γ ⊢ iQ ≈ iP2]]$, and hence $[[nf(iQ) = nf(iP2)]]$ by \cref{lemma:subt-equiv-algorithmization},
    \end{itemize}
    which implies $[[nf(iP1) = nf(iP2)]]$.
    It means that the merge of $[[usEntry1]]$ and $[[usEntry2]]$ is defined by \ruleref{\ottdruleSMEPEqEqLabel}:
    $[[Γ ⊢ (pua :≈ iP1) & (pua :≈ iP2) = (pua :≈ iP1)]]$.

    To show that $[[Γ ⊢ (pua :≈ iQ) ⇒ (pua :≈ iP1)]]$, we apply \ruleref{\ottdruleSImpEPEqEqLabel}:
    $[[Γ ⊢ iP1 ≈ iQ]]$ is obtained from $[[Γ ⊢ iQ ≈ iP1]]$ by symmetry of equivalence.
\end{proof}

\begin{lemma} [Completeness of Unification Solution Merge] 
    \label{lemma:unif-sol-merge-completeness}
    Suppose that $[[us1 : Θ|varset1]]$ and $[[us2 : Θ|varset2]]$ 
    are equivalence-only unification solutions
    and there exists $[[us]]$
    such that $[[Θ ⊢ us ⇒ us1]]$ and $[[ Θ ⊢ us ⇒ us2]]$.
    Then $[[us1 & us2]]$ is defined and 
    $[[ Θ ⊢ us ⇒ us1 & us2 ]]$.
\end{lemma}
\begin{proof}
    By  definition, $[[us1 & us2]]$ is a union of entries of $[[us1]]$, which do not have matching entries in $[[us2]]$,
    entries of $[[us2]]$, which do not have matching entries in $[[us1]]$, and the merge of matching entries.
    It is clear that the first two components of this union exist. Let us show that the third component exists, 
    i.e. that for every two entries $[[usEntry1]] \in [[us1]]$ and $[[usEntry2]] \in [[us2]]$
    restricting the same variable $[[α̂±]]$, $[[usEntry1 & usEntry2]]$ is defined.

    $[[Θ|varset1 ⊢ us ⇒ us1]]$ means that there exists $[[usEntry]] \in [[us]]$ restricting $[[α̂±]]$ such that
    $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1]]$. Similarly, since there can be only one entry in $[[usEntry]]$ restricting 
    the same variable$[[α̂±]]$, $[[Θ2 ⊢ us ⇒ us2]]$ means $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry2]]$.
    Then by \cref{lemma:unif-sol-ent-merge-soundness}, $[[usEntry1 & usEntry2]]$ is defined and 
    $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1 & usEntry2]]$.

    Let us show: $[[ Θ ⊢ us ⇒ us1 & us2 ]]$, i.e. for every $[[usEntry']] \in [[us1 & us2]]$ restricting 
    $[[α̂±]]$, there exists $[[usEntry]] \in [[us]]$ such that $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry']]$.
    Let us consider three cases:
    \begin{caseof}
        \item $[[usEntry']]$ is from $[[us1]]$ and does not have a matching entry in $[[us2]]$.
        Then $[[Θ|varset1 ⊢ us ⇒ us1]]$ means that there exists $[[usEntry]] \in [[us]]$ such that
        $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry']]$, as required.
        \item $[[usEntry']]$ is from $[[us2]]$ and does not have a matching entry in $[[us1]]$.
        This case is symmetric to the previous one.
        \item $[[usEntry']]$ is the merge of two matching entries $[[usEntry1]] \in [[us1]]$
        and $[[usEntry2]] \in [[us2]]$ restricting $[[α̂±]]$.
        Then as noted previously, $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1 & usEntry2]] = [[usEntry']]$.
    \end{caseof}
\end{proof}

\begin{lemma} [Completeness of Solution Entry Merge]
    \label{lemma:entry-merge-completeness}
    For a fixed context $[[Γ]]$,
    suppose that  $[[Γ ⊢ usEntry1]]$ and $[[Γ ⊢ usEntry2]]$.
    Suppose there exists an entry $[[Γ ⊢ usEntry]]$ such that $[[Γ ⊢ usEntry ⇒ usEntry1]]$ and $[[Γ ⊢ usEntry ⇒ usEntry2]]$.
    Then $[[usEntry1 & usEntry2]]$ is defined and
    $[[Γ ⊢ usEntry ⇒ usEntry1 & usEntry2]]$.
\end{lemma}
\begin{proof}
    Notice that $[[usEntry1]]$, $[[usEntry2]]$, and $[[usEntry]]$
    restrict the same variable and have the same polarity. If the polarity is negative,
    then the proof is analogous to the proof of \cref{lemma:unif-entry-merge-completeness}.

    Let us consider the shape of $[[usEntry]]$.
    \begin{caseof}
        \item $[[usEntry]]$ is $[[pua :≥ iP]]$.
        \label{case:entry-merge-completeness-sup-sup}
         Then the only rule inferring 
        $[[Γ ⊢ usEntry ⇒ usEntryi]]$ is \ruleref{\ottdruleSImpESupSupLabel}.
        This way, $[[usEntry1]]$ is $[[pua :≥ iP1]]$ and $[[usEntry2]]$ is $[[pua :≥ iP2]]$,
        and $[[iP]]$ is a common supertype of $[[iP1]]$ and $[[iP2]]$, i.e. $[[Γ⊢ iP ≥ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$.
        Then by \cref{lemma:lub-completeness}, $[[iP1 ∨ iP2]]$ is defined and $[[Γ ⊢ iP ≥ iP1 ∨ iP2]]$.
        Then $[[usEntry1 & usEntry2]]$, that is $[[pua :≥ iP1 & pua :≥ iP1]]$, is defined as 
        $[[pua :≥ iP1 ∨ iP2]]$ by \ruleref{\ottdruleSMESupSupLabel}.
        Moreover, $[[Γ ⊢ usEntry ⇒ usEntry1 & usEntry2]]$, that is $[[Γ ⊢ pua :≥ iP ⇒ pua :≥ iP1 ∨ iP2]]$,
        holds by \ruleref{\ottdruleSImpESupSupLabel}.

        \item $[[usEntry]]$ is $[[pua :≈ iP]]$. Then let us consider the rules inferring
        $[[Γ ⊢ usEntry ⇒ usEntry1]]$ and $[[Γ ⊢ usEntry ⇒ usEntry2]]$ respectively:
        \begin{caseof}
            \item \ruleref{\ottdruleSImpEPEqEqLabel} and \ruleref{\ottdruleSImpEPEqEqLabel}.
            Then $[[usEntry1]]$ is $[[pua :≈ iP1]]$, $[[usEntry2]]$ is $[[pua :≈ iP2]]$, 
            where $[[Γ ⊢ iP ≈ iP1]]$ and $[[Γ ⊢ iP ≈ iP2]]$.
            Then $[[usEntry1 & usEntry2]]$ is defined as $[[pua :≈ iP1]]$ by \ruleref{\ottdruleSMEPEqEqLabel}, 
            and $[[Γ ⊢ pua :≈ iP ⇒ pua :≈ iP1]]$ holds by \ruleref{\ottdruleSImpEPEqEqLabel},
            since $[[Γ ⊢ iP ≈ iP1]]$.

            \item \ruleref{\ottdruleSImpEPEqEqLabel} and \ruleref{\ottdruleSImpEEqSupLabel}.
            Then $[[usEntry1]]$ is $[[pua :≈ iP1]]$, $[[usEntry2]]$ is $[[pua :≥ iP2]]$, 
            where $[[Γ ⊢ iP ≈ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$.

            Then by transitivity of subtyping (\cref{corollary:subtyping-transitivity}),
            $[[Γ ⊢ iP1 ≥ iP2]]$. The completeness of positive subtyping algorithm 
            (\cref{lemma:pos-subt-completeness}) implies that
            $[[Γ;· ⊨ uP1 ≥ iP2 ⫤ us']]$.
            Then $[[usEntry1 & usEntry2]]$ is defined as $[[pua :≈ iP1]]$ by \ruleref{\ottdruleSMEPEqEqLabel}.
            Moreover, $[[Γ ⊢ pua :≈ iP ⇒ pua :≈ iP1]]$ holds by \ruleref{\ottdruleSImpEPEqEqLabel}, 
            as required.

            \item \ruleref{\ottdruleSImpEEqSupLabel} and \ruleref{\ottdruleSImpEPEqEqLabel}.
            This case is symmetric to the previous one.

            \item \ruleref{\ottdruleSImpEEqSupLabel} and \ruleref{\ottdruleSImpEEqSupLabel}.
            Then $[[usEntry1]]$ is $[[pua :≥ iP1]]$, $[[usEntry2]]$ is $[[pua :≥ iP2]]$, 
            where $[[Γ ⊢ iP ≥ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$.
            And similarly to \cref{case:entry-merge-completeness-sup-sup},
            $[[usEntry1 & usEntry2]]$ is defined as $[[pua :≥ iP1 ∨ iP2]]$.
            Moreover, $[[Γ ⊢ pua :≈ iP ⇒ pua :≥ iP1 ∨ iP2]]$ holds by \ruleref{\ottdruleSImpEEqSupLabel},
            since $[[Γ ⊢ iP ≥ iP1 ∨ iP2]]$ by completeness of the least upper bound algorithm (\cref{lemma:lub-completeness}).
        \end{caseof}
    \end{caseof}
\end{proof}

\begin{lemma} [Completeness of Solution Merge] 
    \label{lemma:merge-completeness}
    Suppose that $[[us1 : Θ|varset1]]$ and $[[us2 : Θ|varset2]]$
    and there exists $[[us]]$
    such that $[[Θ ⊢ us ⇒ us1]]$ and $[[Θ ⊢ us ⇒ us2]]$.
    Then $[[us1 & us2]]$ is defined and 
    $[[ Θ ⊢ us ⇒ us1 & us2]]$.
\end{lemma}
\begin{proof}
    The proof is analogous to the proof of \cref{lemma:unif-sol-merge-completeness}.
    The only difference is that we use \cref{lemma:entry-merge-completeness} instead of
    \cref{lemma:unif-entry-merge-completeness}.
\end{proof}
